---
title: "MaxAsymetric"
author: "Maxwell Cornellier"
date: "12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(geosphere)
library(ggmap)
library(lubridate)
```


```{r}
masterdata <- read.csv("new_MASTER_01_data.csv")
#sample_01_data <- read.csv("sample_01_data.csv")
#masterdata$newStartTime <- sample_01_data$starttime
#masterdata$newStopTime <- sample_01_data$stoptime
#write.csv(masterdata,"new_MASTER_01_data.csv")
  
summary(masterdata)
str(masterdata)

#is the date data gone?
#summary(read.csv("sample_01_data.csv"))
#convert columns to factors as needed

masterdata$bikeid <- as.factor(masterdata$bikeid)
masterdata$gender <- as.factor(masterdata$gender)

masterdata$gender <- as.factor(ifelse(masterdata$gender == "0", "Unknown", ifelse(masterdata$gender == "1", "Male", "Female")))
masterdata$X <- NULL
masterdata$starttime <- NULL
masterdata$stoptime <- NULL

masterdata$newStartTime = as.POSIXct(strptime(masterdata$newStartTime, "%Y-%m-%d %H:%M:%S"))
masterdata$newStopTime = as.POSIXct(strptime(masterdata$newStopTime, "%Y-%m-%d %H:%M:%S"))

masterdata$newStartDate <- as.Date(masterdata$newStartTime)
masterdata$newStopDate <- as.Date(masterdata$newStopTime)


#distance
masterstart <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterstart$startlong <- as.numeric(masterdata$start.station.longitude)
masterstart$startlat <- as.numeric(masterdata$start.station.latitude)

masterend <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterend$endlong <- masterdata$end.station.longitude
masterend$endlat <- masterdata$end.station.latitude

masterdata$distanceH <- distHaversine(masterstart, masterend, r=6378137)

masterend <- NULL
masterstart <- NULL

masterdata$startMonth <- month(masterdata$newStartDate)
masterdata$stopMonth <- month(masterdata$newStopDate)
masterdata$startMonthFactor <- as.factor(month(masterdata$newStartDate))
masterdata$stopMonthFactor <- as.factor(month(masterdata$newStopDate))

#Create Time of Day Factor 

masterdata$timeOfDay <- ifelse(hour(masterdata$newStartTime) >= 0 & hour(masterdata$newStartTime) < 12, "morning", ifelse(hour(masterdata$newStartTime) >= 12 & hour(masterdata$newStartTime) <=24, "afternoon","night"))


masterdata$timeOfDay <- as.factor(masterdata$timeOfDay)

# Create data frame for AM and one for PM

masterAM <- filter(masterdata, masterdata$timeOfDay == "morning")
masterPM <- filter(masterdata,masterdata$timeOfDay == "afternoon")



str(masterdata)
summary(masterdata)






```


# What Stations Gain or Lose Bikes throughout the day ? 

```{r}

# Unique Departures and Arrivals for Each Citi Bike Station
StationStartsAM <- as.data.frame(table(masterAM$start.station.name))
StationEndsAM <- as.data.frame(table(masterAM$end.station.name))

StationDataAM <- data.frame(masterAM$start.station.name)
StationDataAM <- unique(StationDataAM)
StationDataAM$numStarts <- StationStartsAM$Freq[match(StationDataAM$masterAM.start.station.name, StationStartsAM$Var1)]
StationDataAM$numEnds <- StationEndsAM$Freq[match(StationDataAM$masterAM.start.station.name, StationEndsAM$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataAM$difference <- StationDataAM$numEnds - StationDataAM$numStarts
StationDataAM <- arrange(StationDataAM, desc(difference))
StationDataAM <- na.omit(StationDataAM)

# Top 10 stations that gain bikes throughout the morning
TopTenSurplusAM <- head(StationDataAM, 10)

# Top 10 stations that lose bikes throughout the morning 

TopTenDeficitAM <-tail(StationDataAM, 10)
TopTenDeficitAM <- arrange(TopTenDeficitAM, difference)


#Replicate for PM times 


# Unique Departures and Arrivals for Each Citi Bike Station
StationStartsPM <- as.data.frame(table(masterPM$start.station.name))
StationEndsPM <- as.data.frame(table(masterPM$end.station.name))

StationDataPM <- data.frame(masterPM$start.station.name)
StationDataPM <- unique(StationDataPM)
StationDataPM$numStarts <- StationStartsPM$Freq[match(StationDataPM$masterPM.start.station.name, StationStartsPM$Var1)]
StationDataPM$numEnds <- StationEndsPM$Freq[match(StationDataPM$masterPM.start.station.name, StationEndsPM$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataPM$difference <- StationDataPM$numEnds - StationDataPM$numStarts
StationDataPM <- arrange(StationDataPM, desc(difference))
StationDataPM <- na.omit(StationDataPM)

# Top 10 stations that gain bikes throughout the morning
TopTenSurplusPM <- head(StationDataPM, 10)

# Top 10 stations that lose bikes throughout the morning 

TopTenDeficitPM <-tail(StationDataPM, 10)
TopTenDeficitPM <- arrange(TopTenDeficitPM, difference)


TopTenSurplusAM
TopTenDeficitAM
TopTenSurplusPM
TopTenDeficitPM
```

### The Top 10 Stations that have more arrivals than departures in the mornining (gain bikes) are 

### The Top 10 Stations that have more departures than arrivals in the mornining (lose bikes) are 


### Visualize Top Tens 

```{r}

# Top Ten gains in the morning , add chart title 

TenSurplusMorning <- ggplot(TopTenSurplusAM, aes(reorder(masterAM.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusMorning
```
## Analysis: this chart shows a high level of asymmetry among the bike stations with station "Broadwat & E 22 St" having 384 more arrivals than departures in the morning throughout the year. 

```{r}
# Top Ten loses in the morning , add chart title 

TenDeficitMorning <- ggplot(TopTenDeficitAM, aes(reorder(masterAM.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitMorning


```
## Analysis: Asymmetric traffic patterns are also evident in the chart above, with station "8 Ave & W 31 St" experiencing 449 more departures than arrivals in the morning throughout the year. 


```{r}
# Top Ten gains in the afternoon , add chart title 

TenSurplusAfternoon<- ggplot(TopTenSurplusPM, aes(reorder(masterPM.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusAfternoon
```
### Analysis: The evidence of aymmetric traffic is also present in this chart, with station " 8 Ave & W 31 St " leading with 418 more arrivals than departures. Some symmetry between the two time periods is seen here with a loss of 449 bikes at this same station in the morning time period throughout the year. In fact, 8 of the top 10 stations that lose bikes in the morning period are present here gaining bikes. 


```{r}
# Top Ten loses in the afternoon , add chart title 

TenDeficitAfternoon <- ggplot(TopTenDeficitPM, aes(reorder(masterPM.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitAfternoon



```

## Analysis:

### In the chart for the top ten stations that experience losses all ten of the stations that experience overflow of bikes in the morning time perion are present here. 

### What user types are using these top asymmetric stations?

```{r}

CountCustomer <- nrow(masterAM[masterAM$usertype == "Customer" & masterAM$start.station.name == "Broadway & E 22 St",])
CountSubscriber <- nrow(masterAM[masterAM$usertype == "Subscriber" & masterAM$start.station.name == "Broadway & E 22 St",])

#able to find customer type for start, need to figure out for end 



```

## Analysis 

### For stations that are asymmetric, users tend to be "subscribers" as opposed to "customers". At the station "Broadway & E 22 St" of the 239 users who started trips there `r CountSubscriber` are subscribers while only `r CountCustomer`


## What impact does day of the week have on this 



#Map 

```{r}
# Add longitude and latitude to the dataset

startlatitude <- c(40.7403432,40.72019521,40.75510267, 40.76132983,40.75992262,40.72243797,40.7643971,40.74096374, 40.75724568, 40.70463334)

TopTenSurplusAM$startlatitude <- startlatitude

startlongitude <- c(-73.98955109,-74.01030064,-73.97498696
,-73.97982001, -73.97648516, -74.00566443
, -73.97371465, -73.98602213, -73.97805914
, -74.01361706)

TopTenSurplusAM$startlongitude <- startlongitude

```





#Map Showing top ten surplus AM

```{r}

#Map TopTenSurplus AM


register_google(key = "AIzaSyDr6TG5wIRo6iXXvRbE0rV3n2EPx1jApRc")

## get station info
station.info <- TopTenSurplusAM %>%
  group_by(masterAM.start.station.name) %>%
  summarise(lat=as.numeric(startlatitude),
            long=as.numeric(startlongitude),
            difference = difference)


## get map and plot station locations 
newyork.map <- get_map(location= 'Lower Manhattan, New York', 
                       maptype='roadmap', color='bw',source='google',zoom=12)

ggmap(newyork.map) + 
  geom_point(data=station.info,aes(x=long,y=lat,color= difference),size=5,alpha=0.75)+
  scale_colour_gradient(high="red",low='green')+ 
  theme(axis.ticks = element_blank(),axis.text = element_blank())+
  xlab('')+ylab('')
```



