---
title: "MaxAsymetric"
author: "Maxwell Cornellier"
date: "12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(geosphere)
library(ggmap)
library(lubridate)
```


```{r}
masterdata <- read.csv("new_MASTER_01_data.csv")
#sample_01_data <- read.csv("sample_01_data.csv")
#masterdata$newStartTime <- sample_01_data$starttime
#masterdata$newStopTime <- sample_01_data$stoptime
#write.csv(masterdata,"new_MASTER_01_data.csv")


#is the date data gone?
#summary(read.csv("sample_01_data.csv"))
#convert columns to factors as needed

masterdata$bikeid <- as.factor(masterdata$bikeid)
masterdata$gender <- as.factor(masterdata$gender)

masterdata$gender <- as.factor(ifelse(masterdata$gender == "0", "Unknown", ifelse(masterdata$gender == "1", "Male", "Female")))
masterdata$X <- NULL
masterdata$starttime <- NULL
masterdata$stoptime <- NULL

masterdata$newStartTime = as.POSIXct(strptime(masterdata$newStartTime, "%Y-%m-%d %H:%M:%S"))
masterdata$newStopTime = as.POSIXct(strptime(masterdata$newStopTime, "%Y-%m-%d %H:%M:%S"))

masterdata$newStartDate <- as.Date(masterdata$newStartTime)
masterdata$newStopDate <- as.Date(masterdata$newStopTime)


#distance
masterstart <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterstart$startlong <- as.numeric(masterdata$start.station.longitude)
masterstart$startlat <- as.numeric(masterdata$start.station.latitude)

masterend <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterend$endlong <- masterdata$end.station.longitude
masterend$endlat <- masterdata$end.station.latitude

masterdata$distanceH <- distHaversine(masterstart, masterend, r=6378137)

masterend <- NULL
masterstart <- NULL

masterdata$startMonth <- month(masterdata$newStartDate)
masterdata$stopMonth <- month(masterdata$newStopDate)
masterdata$startMonthFactor <- as.factor(month(masterdata$newStartDate))
masterdata$stopMonthFactor <- as.factor(month(masterdata$newStopDate))

#Create Time of Day Factor 

masterdata$timeOfDay <- ifelse(hour(masterdata$newStartTime) >= 0 & hour(masterdata$newStartTime) < 12, "morning", ifelse(hour(masterdata$newStartTime) >= 12 & hour(masterdata$newStartTime) <=24, "afternoon","night"))


masterdata$timeOfDay <- as.factor(masterdata$timeOfDay)
#Day of Week 

masterdata$numWeekday <- as.factor(wday(masterdata$newStartDate))


# Create data frame for AM and one for PM

masterAM <- filter(masterdata, masterdata$timeOfDay == "morning")
masterPM <- filter(masterdata,masterdata$timeOfDay == "afternoon")


# Create subsets for Weekday and Weekend
weekdayData <- subset(masterAM, subset = (masterAM$numWeekday != "7" & masterAM$numWeekday != "1"))
weekendData <- subset(masterAM, subset = (masterAM$numWeekday == "7" | masterAM$numWeekday == "1"))



```


# What Stations Gain or Lose Bikes throughout the day ? 

```{r}

# Unique Departures and Arrivals for Each Citi Bike Station
StationStartsAM <- as.data.frame(table(masterAM$start.station.name))
StationEndsAM <- as.data.frame(table(masterAM$end.station.name))

StationDataAM <- data.frame(masterAM$start.station.name)
StationDataAM <- unique(StationDataAM)
StationDataAM$numStarts <- StationStartsAM$Freq[match(StationDataAM$masterAM.start.station.name, StationStartsAM$Var1)]
StationDataAM$numEnds <- StationEndsAM$Freq[match(StationDataAM$masterAM.start.station.name, StationEndsAM$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataAM$difference <- StationDataAM$numEnds - StationDataAM$numStarts
StationDataAM <- arrange(StationDataAM, desc(difference))
StationDataAM <- na.omit(StationDataAM)

# Top 10 stations that gain bikes throughout the morning

TopTenSurplusAM <- head(StationDataAM, 10)


# Top 10 stations that lose bikes throughout the morning 

TopTenDeficitAM <-tail(StationDataAM, 10)
TopTenDeficitAM <- arrange(TopTenDeficitAM, difference)


#Replicate for PM times 


# Unique Departures and Arrivals for Each Citi Bike Station
StationStartsPM <- as.data.frame(table(masterPM$start.station.name))
StationEndsPM <- as.data.frame(table(masterPM$end.station.name))

StationDataPM <- data.frame(masterPM$start.station.name)
StationDataPM <- unique(StationDataPM)
StationDataPM$numStarts <- StationStartsPM$Freq[match(StationDataPM$masterPM.start.station.name, StationStartsPM$Var1)]
StationDataPM$numEnds <- StationEndsPM$Freq[match(StationDataPM$masterPM.start.station.name, StationEndsPM$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataPM$difference <- StationDataPM$numEnds - StationDataPM$numStarts
StationDataPM <- arrange(StationDataPM, desc(difference))
StationDataPM <- na.omit(StationDataPM)

# Top 10 stations that gain bikes throughout the morning
TopTenSurplusPM <- head(StationDataPM, 10)

# Top 10 stations that lose bikes throughout the morning 

TopTenDeficitPM <-tail(StationDataPM, 10)
TopTenDeficitPM <- arrange(TopTenDeficitPM, difference)


TopTenSurplusAM

TopTenDeficitAM

TopTenSurplusPM

TopTenDeficitPM

```


## Visualize Top Tens 

```{r}

# Top Ten gains in the morning 

TenSurplusMorning <- ggplot(TopTenSurplusAM, aes(reorder(masterAM.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusMorning
```
Analysis: 
this chart shows a high level of asymmetry among the bike stations with station "Broadwat & E 22 St" having 384 more arrivals than departures in the morning throughout the year. 

```{r}
# Top Ten loses in the morning 
TenDeficitMorning <- ggplot(TopTenDeficitAM, aes(reorder(masterAM.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitMorning


```
Analysis: 
Asymmetric traffic patterns are also evident in the chart above, with station "8 Ave & W 31 St" experiencing 449 more departures than arrivals in the morning throughout the year. 


```{r}
# Top Ten gains in the afternoon 

TenSurplusAfternoon<- ggplot(TopTenSurplusPM, aes(reorder(masterPM.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusAfternoon
```

Analysis:
The evidence of asymmetric traffic is also present in this chart, with station " 8 Ave & W 31 St " leading with 418 more arrivals than departures. Some symmetry between the two time periods is seen here with a loss of 449 bikes at this same station in the morning time period throughout the year. In fact, 8 of the top 10 stations that lose bikes in the morning period are present here gaining bikes. 


```{r}
# Top Ten loses in the afternoon 

TenDeficitAfternoon <- ggplot(TopTenDeficitPM, aes(reorder(masterPM.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitAfternoon



```

Analysis:

In the chart for the top ten stations that experience losses all ten of the stations that experience overflow of bikes in the morning time perion are present here. 


## Does Weekend vs. Weekday have an effect on asymmetric traffic? Test Hypothesis that asymmetry is due to work commutes 
For this I compared weekdays vs . weekends for data from the first 12 hours of the data (masterAM)

```{r}

#create counts of data 
# Unique Departures and Arrivals for Each Citi Bike Station for Week 
StationStartsWeek <- as.data.frame(table(weekdayData$start.station.name))
StationEndsWeek <- as.data.frame(table(weekdayData$end.station.name))

StationDataWeek <- data.frame(weekdayData$start.station.name)
StationDataWeek <- unique(StationDataWeek)
StationDataWeek$numStarts <- StationStartsWeek$Freq[match(StationDataWeek$weekdayData.start.station.name, StationStartsWeek$Var1)]
StationDataWeek$numEnds <- StationEndsWeek$Freq[match(StationDataWeek$weekdayData.start.station.name, StationEndsWeek$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataWeek$difference <- StationDataWeek$numEnds - StationDataWeek$numStarts
StationDataWeek <- arrange(StationDataWeek, desc(difference))
StationDataWeek <- na.omit(StationDataWeek)

# Top 10 stations that gain bikes throughout Weekdays

TopTenSurplusWeek <- head(StationDataWeek, 10)


# Top 10 stations that lose bikes throughout Weekend Mornings 

TopTenDeficitWeek <-tail(StationDataWeek, 10)
TopTenDeficitWeek <- arrange(TopTenDeficitWeek, difference)


#create counts of data 
# Unique Departures and Arrivals for Each Citi Bike Station for Weekend  
StationStartsWeekend <- as.data.frame(table(weekendData$start.station.name))
StationEndsWeekend <- as.data.frame(table(weekendData$end.station.name))

StationDataWeekend <- data.frame(weekendData$start.station.name)
StationDataWeekend <- unique(StationDataWeekend)
StationDataWeekend$numStarts <- StationStartsWeekend$Freq[match(StationDataWeekend$weekendData.start.station.name, StationStartsWeekend$Var1)]
StationDataWeekend$numEnds <- StationEndsWeekend$Freq[match(StationDataWeekend$weekendData.start.station.name, StationEndsWeekend$Var1)]


# Compute the difference (Arrivals > Departures)
StationDataWeekend$difference <- StationDataWeekend$numEnds - StationDataWeekend$numStarts
StationDataWeekend <- arrange(StationDataWeekend, desc(difference))
StationDataWeekend <- na.omit(StationDataWeekend)

# Top 10 stations that gain bikes throughout Weekend Mornings 

TopTenSurplusWeekend <- head(StationDataWeekend, 10)


# Top 10 stations that lose bikes throughout Weekend Mornings 

TopTenDeficitWeekend <-tail(StationDataWeekend, 10)
TopTenDeficitWeekend <- arrange(TopTenDeficitWeekend, difference)

```


## Visuals 

## Top Surplus During Week Mornings

```{r}
TenSurplusWeek <- ggplot(TopTenSurplusWeek, aes(reorder(weekdayData.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusWeek
```

## Top Deficit during Week Mornings

```{r}
TenDeficitWeek <- ggplot(TopTenDeficitWeek, aes(reorder(weekdayData.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitWeek
```
### Top Surplus During Weekend Mornings

```{r}
TenSurplusWeekend <- ggplot(TopTenSurplusWeekend, aes(reorder(weekendData.start.station.name, - difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenSurplusWeekend
```
## Top Deficit during Weekend Mornings

```{r}
TenDeficitWeekend <- ggplot(TopTenDeficitWeekend, aes(reorder(weekendData.start.station.name,  difference), difference)) +
    geom_col() +
    scale_x_discrete(label = function(x) stringr::str_trunc(x, 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
TenDeficitWeekend
```
Analysis:
Between the top deficit stations in the morning of the weekends and weekdays there is only one overlapping station: E 13th and Avenue A., further the highest deficit on weekends is around 40 (average 20 a day [40/2]) whereas during the week the highest is approximately 450(average 90 a day [450/5]). Therefore this analysis shows that there is much more asymmetry during the week as opposed to the weekend, when less people commute for work. This pattern holds true for top surplus stations in the mornings of the weekend and weekdays with no overlapping stations,and the top station during weekdays with a surplus of 375 (average 75 a day [375/5])compared to a top surplus of 55 on the weekend (average 28 a day [55/2]). In summary, this supports the hypothesis that asymmetry is mainly caused by commutes to work during the week. 


## What user types are using these top asymmetric stations?

```{r}

CountCustomerBroadway <- nrow(masterAM[masterAM$usertype == "Customer" & masterAM$start.station.name == "Broadway & E 22 St",])
CountSubscriberBroadway <- nrow(masterAM[masterAM$usertype == "Subscriber" & masterAM$start.station.name == "Broadway & E 22 St",])
CountCustomerNMoore <- nrow(masterPM[masterPM$usertype == "Customer" & masterPM$start.station.name == "North Moore St & Greenwich St",])
CountSubscriberNMoore <- nrow(masterPM[masterPM$usertype == "Subscriber" & masterPM$start.station.name == "North Moore St & Greenwich St",])



```

Analysis:
For stations that are asymmetric, users tend to be "subscribers" as opposed to "customers". At the station "Broadway & E 22 St" of the 239 users who started trips there `r CountSubscriberBroadway` are subscribers while only `r CountCustomerBroadway`. For another station "North Moore St & Greenwich" there was only `r CountCustomerNMoore`customers while there was `r CountSubscriberNMoore` Subscribers. 

#Map 

```{r}
# Add longitude and latitude to the dataset

startlatitude <- c(40.7403432,40.72019521,40.75510267, 40.76132983,40.75992262,40.72243797,40.7643971,40.74096374, 40.75724568, 40.70463334)

TopTenSurplusAM$startlatitude <- startlatitude

startlongitude <- c(-73.98955109,-74.01030064,-73.97498696
,-73.97982001, -73.97648516, -74.00566443
, -73.97371465, -73.98602213, -73.97805914
, -74.01361706)

TopTenSurplusAM$startlongitude <- startlongitude

```


##Map Showing top ten surplus AM


```{r}

#Map TopTenSurplus AM


register_google(key = "AIzaSyDr6TG5wIRo6iXXvRbE0rV3n2EPx1jApRc")

## get station info
station.info <- TopTenSurplusAM %>%
  group_by(masterAM.start.station.name) %>%
  summarise(lat=as.numeric(startlatitude),
            long=as.numeric(startlongitude),
            difference = difference)


## get map and plot station locations 
newyork.map <- get_map(location= 'Lower Manhattan, New York', 
                       maptype='roadmap', color='bw',source='google',zoom=12)

ggmap(newyork.map) + 
  geom_point(data=station.info,aes(x=long,y=lat,color= difference),size=5,alpha=0.75)+
  scale_colour_gradient(high="red",low='green')+ 
  theme(axis.ticks = element_blank(),axis.text = element_blank())+
  xlab('')+ylab('')
```
### Analysis 

As seen in the geographic groupings of the top asymmetric stations most of them are located in Lower Manhattan, specifically in areas like Midtown where there are many jobs which is true for pretty much all of these areas. Although this map only shows the top stations that gain bikes in the morning, this pattern is true for the asymmetric stations due to the overlap from the realtionship between the stations. 


