---
title: "NamanPlayground"
author: "Naman Gupta"
date: "12/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
```

```{r}
masterdata <- read.csv("MASTERsample_01_data.csv")
summary(masterdata)
str(masterdata)
#is the date data gone?

#convert columns to factors as needed

masterdata$bikeid <- as.factor(masterdata$bikeid)
masterdata$gender <- as.factor(masterdata$gender)
masterdata$gender <- as.factor(ifelse(masterdata$gender == "0", "Unknown", ifelse(masterdata$gender == "1", "Male", "Female")))
masterdata$X <- NULL
summary(masterdata)
str(masterdata)
```
#Plot #1: A scatter plot with weight on x-axis, hip on y-axis, and gender as the color of points. What do you see - write a quick summary of insights from this plot.
```{r}
#average precip per month
#facet wrap month, against count of number of rides as y variable. x = average precip, colour = gender
#categorize birth year into age groups. Facet wrap or colour by age groups
#rain vs trip length (end time - start time)
#bar plot of busy/not-busy stations on rainy days


#randomly try the same plot across a few select dates


scatter <- ggplot(data=masterdata, aes(x=PRCP, y=, colour=gender)) + geom_point(size=2, alpha=1)
scatter
```
>The variables weight and hip are strongly positively correlated. Generally, females have higher hip sizes than males with relation to weight.


#Plot#2: Experiement with the scatter plot you just built. Try these things:size,alpha,shape,Adding another layer of geom_smooth
```{r}
baseplot1 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1) + geom_smooth()
baseplot1

baseplot2 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1) + geom_smooth(method = "lm")
baseplot2

baseplot3 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=50, alpha=1) + geom_smooth(method = "lm")
baseplot3

baseplot4 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1)
baseplot4

baseplot5 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1)+ geom_smooth(method = "gam")
baseplot5

baseplot6 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1)+ geom_smooth(method = "glm")
baseplot6

baseplot7 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1)+ geom_smooth(method = "loess")
baseplot7

baseplot8 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1)+ geom_smooth(method = "lm") + geom_smooth()
baseplot8

baseplot9 <- ggplot(data=diabetes, aes(x=weight, y=hip, colour=gender)) + geom_point(size=2, alpha=1, shape = 3) + geom_smooth(method = "lm") + geom_smooth()
baseplot9
```
>We experimented with the various things listed on the assignment. We found that adding an extra geom_smooth() results in an additional line on the graphic. After a certain size level, we found that it does not make sense to have very large sized points, as they appear to simply merge into each other.

#Plot#3: Build a Box Plot with gender on x-axis and chol on y-axis
```{r}
boxplot <- ggplot(data=diabetes, aes(x=gender, y=chol, colour=gender)) + geom_boxplot()
boxplot
```
>This shows us the distribution of the chol variable separated by the gender variable. It appears that both genders have very similar distributions.

#Plot#4: Change the Box Plot to a Violin Plot
```{r}
violin <- ggplot(data=diabetes, aes(x=gender, y=chol, colour=gender)) + geom_violin()
violin
```
>This shows us that Cholesterol distribution for females and males is centered around 200, with females having higher cholesterol on average and men having lower cholesterol.

#Plot#5: Built a histogram with glyhb as the variable plotted
```{r}
histogram <- ggplot(data=diabetes, aes(x=glyhb)) + geom_histogram()
histogram
```
>In this dataset, we see the highest count of people with glyhb levrls centering around a value of 5. The number of people with higher levels drastically decreases as glyhb values increase.

#Plot#6: Change the histogram to a density plot
```{r}
density <- ggplot(data=diabetes, aes(x=glyhb)) + geom_density()
density
```
>Here we can see as the hemoglobin's density is heavy around a value of 5 and then drastically decreases as the hemoglobin increases.


#Plot#7: Build a Bar Plot with variable frame on x-axis
```{r}
barplot <- ggplot(data=diabetes, aes(x=frame)) + geom_bar()
barplot
```
