---
title: "TempAnalysis_Dawson"
author: "Dawson Wang"
date: "12/2/2020"
output: 
  html_document:
      number_sections: true
      toc: true
      toc_float: 
        toc_collapsed: true
      toc_depth: 3
      theme: lumen
---

# Initialize

## Packages
```{r}
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gganimate)
library(gifski)
library(png)
library(data.table)
library(geosphere)
library(leaflet)
```

## Check File
```{r}
master <- read.csv("new_MASTER_01_data.csv")
```

## Clean Data
```{r}
master$X <- NULL
master$start.station.id <- as.factor(master$start.station.id)
master$start.station.name <- as.factor(master$start.station.name)
master$end.station.id <- as.factor(master$end.station.id)
master$end.station.name <- as.factor(master$end.station.name)
master$bikeid <- as.factor(master$bikeid)
master$usertype <- as.factor(master$usertype)
master$gender <- as.factor(master$gender)
master <- rename(master, startTime = newStartTime, stopTime = newStopTime)
master$startTime <- as.POSIXct(strptime(master$startTime, "%Y-%m-%d %H:%M:%S"))
master$stopTime <- as.POSIXct(strptime(master$stopTime, "%Y-%m-%d %H:%M:%S"))
master$startDate <- as.Date(master$startTime)
master$stopDate <- as.Date(master$stopTime)
master$distMeters <- distHaversine(cbind(master$start.station.latitude, master$start.station.longitude), cbind(master$end.station.latitude, master$end.station.longitude))
master$ageGroup <- as.factor(ifelse(master$birth.year >= 2000, "GenZ", ifelse(master$birth.year >= 1981, "Millennial", ifelse(master$birth.year >= 1965, "GenX", ifelse(master$birth.year >= 1946, "Boomer", ifelse(master$birth.year >= 1928, "Silent", "VeryOld"))))))
master$ageGroup <- factor(master$ageGroup, levels = c("GenZ", "Millennial", "GenX", "Boomer", "Silent", "VeryOld"))
master$startMonth <- month(master$startDate)
master$stopMonth <- month(master$stopDate)
master$seasonStart <- as.factor(ifelse(master$startMonth >= 3 & master$startMonth <= 5, "Spring", ifelse(master$startMonth >= 6 & master$startMonth <= 8, "Summer", ifelse(master$startMonth >= 9 & master$startMonth <= 11, "Fall", "Winter"))))
master$seasonStart <- factor(master$startMonth, levels = c("Spring", "Summer", "Fall", "Winter"))
master$numWeekday <- as.factor(wday(master$startDate))
#Defining rush hour as 6-10AM and 4-8PM
master$rushHour <- as.factor(ifelse(master$numWeekday == 1 | master$numWeekday == 7, "No", ifelse(hour(master$startTime) < 6 | hour(master$startTime) > 10 & hour(master$startTime) < 16 | hour(master$startTime) > 20, "No", "Yes")))
master <- rename(master, maxTemp = TMAX, minTemp = TMIN)
master$weekNum <- as.numeric(strftime(master$startDate, format = "%V"))
master$speedMetersperMin <- master$distMeters / master$tripduration
master <- rename(master, avgTemp = TAVG)
master$tempFeel <- as.factor(ifelse(master$maxTemp < 40, "Frigid", ifelse(master$maxTemp < 58, "Cold", ifelse(master$maxTemp < 65, "Cool", ifelse(master$maxTemp < 75, "Warm", ifelse(master$maxTemp < 95, "Hot", "Blazing"))))))
master$tempFeel <- factor(master$tempFeel, levels = c("Frigid", "Cold", "Cool", "Warm", "Hot", "Blazing"))

```

## Updated Data Frame
```{r}
str(master)
summary(master)
```

# Data Exploration

## Average Daily Rides
```{r}
#By Day of Week
master %>%
  group_by(startDate, numWeekday) %>%
  summarise(
    numRides = mean(n())
  ) %>%
  ggplot(aes(x = numWeekday, y = numRides)) + geom_boxplot() + labs(x = "Weekday (1 = Sunday, 7 = Saturday)", y = "Average Number of CitiBike Rides")
#By Maximum Temperature
master %>%
  group_by(maxTemp) %>%
  summarise(
    numRides = mean(n())
  ) %>%
  ggplot(aes(x = maxTemp, y = numRides)) + geom_point() + labs(x = "Maximum Temperature (F)", y = "Average Number of CitiBike Rides")
#By Minimum Temperature
master %>%
  group_by(minTemp) %>%
  summarise(
    numRides = mean(n())
  ) %>%
  ggplot(aes(x = minTemp, y = numRides)) + geom_point() + labs(x = "Minimum Temperature (F)", y = "Average Number of CitiBike Rides")
#By Day of Week by Temperature (Not super meaningful - keeping this here as a template)
master %>%
  group_by(weekNum, numWeekday) %>%
  summarise(
    numRides = mean(n()),
    temp = maxTemp
  ) %>%
  ggplot(aes(x = temp, y = numRides)) + geom_boxplot() + transition_time(weekNum) + labs(title = "Week Number: {frame_time}")
```

## Ride Duration
```{r}
#By Maximum Temperature By Gender
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = gender)) + geom_point()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = gender)) + geom_smooth()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = gender)) + geom_violin()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = gender)) + geom_boxplot()
#By Maximum Temperature By User Type
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = usertype)) + geom_point()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = usertype)) + geom_smooth()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = usertype)) + geom_violin()
ggplot(data = master[master$tripduration < 10000,], aes(x = maxTemp, y = tripduration, colour = usertype)) + geom_boxplot()
#By Minimum Temperature By Gender
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = gender)) + geom_point()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = gender)) + geom_smooth()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = gender)) + geom_violin()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = gender)) + geom_boxplot()
#By Minimum Temperature By User Type
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = usertype)) + geom_point()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = usertype)) + geom_smooth()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = usertype)) + geom_violin()
ggplot(data = master[master$tripduration < 10000,], aes(x = minTemp, y = tripduration, colour = usertype)) + geom_boxplot()
```

## Ride Distance
```{r}
#Ride Distance by Maximum Temperature by Gender 
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = gender)) + geom_point()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = gender)) + geom_smooth()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = gender)) + geom_violin()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = gender)) + geom_boxplot()
#Ride Distance by Maximum Temperature by User Type
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = usertype)) + geom_point()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = usertype)) + geom_smooth()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = usertype)) + geom_violin()
ggplot(data = master, aes(x = maxTemp, y = distMeters, colour = usertype)) + geom_boxplot()
#Ride Distance by Minimum Temperature by Gender
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = gender)) + geom_point()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = gender)) + geom_smooth()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = gender)) + geom_violin()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = gender)) + geom_boxplot()
#Ride Distance by Minimum Temperature by User Type
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = usertype)) + geom_point()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = usertype)) + geom_smooth()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = usertype)) + geom_violin()
ggplot(data = master, aes(x = minTemp, y = distMeters, colour = usertype)) + geom_boxplot()
```

## Speed
```{r}
#Speed by Maximum Temperature by Gender
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = gender)) + geom_point()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = gender)) + geom_smooth()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = gender)) + geom_violin()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = gender)) + geom_boxplot()
#Speed by Maximum Temperature by User Type
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = usertype)) + geom_point()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = usertype)) + geom_smooth()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = usertype)) + geom_violin()
ggplot(data = master, aes(x = maxTemp, y = speedMetersperMin, colour = usertype)) + geom_boxplot()
#Speed by Minimum Temperature by Gender
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = gender)) + geom_point()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = gender)) + geom_smooth()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = gender)) + geom_violin()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = gender)) + geom_boxplot()
#Speed by Minimum Temperature by User Type
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = usertype)) + geom_point()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = usertype)) + geom_smooth()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = usertype)) + geom_violin()
ggplot(data = master, aes(x = minTemp, y = speedMetersperMin, colour = usertype)) + geom_boxplot()
```

##Station Utilization
```{r}
master %>%
  group_by(tempFeel, start.station.id) %>%
  summarise(
    numRides = mean(n()),
  ) %>%
  arrange(desc(numRides)) %>%
  slice(1:5) %>%
  ggplot(., aes(x = tempFeel, y = numRides, colour = start.station.id )) + geom_boxplot()
```


# Non-Temperature Analysis Ideas
## Age Outliers/Anomalies
## Count of rides by men/women; men could ride bikes more times than women, or there may be more men subscribers than women - we can't tell
## How Bikes Should Be Distributed