---
title: "TO404_FinalProject"
author: "Brian Zhao"
date: "12/4/2020"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_float: 
      toc_collapsed: true
    toc_depth: 3
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# TO 404 Final Project

## Libraries
```{r}
library(dplyr)
library(ggplot2)
library(geosphere)
library(data.table)
```

## Read Data
```{r, cache=TRUE, eval = FALSE}
# Read Original CitiBike Data
sampleData <- read.csv("SampleCitiBike.csv")
# Create 1% Random Sample
sampleData_01 <- sample_frac(data, 0.01)
# Write to New CSV
write.csv(sampleData_01, 'sample_01_data.csv')
# Weather Data Manually Appended in new_MASTER_01_data.csv
```

```{r}
masterdata <- read.csv("new_MASTER_01_data.csv")
```

## Clean Data
```{r}
# Explore Data
str(masterdata)
summary(masterdata)

# Remove Identifier Column
masterdata$X <- NULL

# Assign Factors
masterdata$bikeid <- as.factor(masterdata$bikeid)
masterdata$gender <- as.factor(masterdata$gender)
masterdata$gender <- as.factor(ifelse(masterdata$gender == "0", "Unknown", ifelse(masterdata$gender == "1", "Male", "Female")))

# Distance
masterstart <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterstart$startlong <- as.numeric(masterdata$start.station.longitude)
masterstart$startlat <- as.numeric(masterdata$start.station.latitude)

masterend <- as.data.frame(matrix(nrow = 205517, ncol = 0))
masterend$endlong <- masterdata$end.station.longitude
masterend$endlat <- masterdata$end.station.latitude

masterdata$distanceH <- distHaversine(masterstart, masterend, r=6378137)

# Speed
masterdata$speedMetersperMin <- masterdata$distanceH / masterdata$tripduration

# Date
masterdata$newStartTime = as.POSIXct(strptime(masterdata$newStartTime, "%Y-%m-%d %H:%M:%S"))
masterdata$newStopTime = as.POSIXct(strptime(masterdata$newStopTime, "%Y-%m-%d %H:%M:%S"))
masterdata$startDate <- as.Date(masterdata$newStartTime)

# Snow
masterdata$roundedSNOW <- floor(masterdata$SNOW)

# Check Data
str(masterdata)
summary(masterdata)
```

# Weather

## Snow
*Effects of Snow on:*

- Ride Distance (Customer vs. Subscriber)
```{r}
ggplot(data=masterdata[masterdata$distanceH < 10000,], aes(x=SNOW, y=distanceH, colour=usertype)) + geom_smooth() + labs(title = "Effects of Snow on Citi Bike Ride Distance", x = "Snow Depth (inches)", y = "Distance (meters)")

ggplot(data=masterdata[masterdata$distanceH < 10000,], aes(x=SNOW, y=distanceH, colour=usertype)) + geom_point(alpha = 0.5) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Ride Distance", x = "Snow Depth (inches)", y = "Distance (meters)")

ggplot(data=masterdata[masterdata$distanceH < 10000,], aes(x=roundedSNOW, y=distanceH, colour=usertype)) + geom_point(alpha = 0.5) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Ride Distance", x = "Snow Depth (rounded inches)", y = "Distance (meters)")
```

We can see the trend that customers generally ride a consistent distance regardless of snow, while subscribers tend to travel shorter distances when there is snow. This is likely due to the fact that subscribers or frequent users are more likely to be locals to NYC and use other forms of transportation instead (i.e. subway).

However, if we also factor in the amount of rides that are happening, we can see that subscribers make up a larger percentage of the total rides compared to customers whenever there is snow on the ground. Subscribers who are more likely to use a CitiBike than a regular customer in the snow (but will also ride a shorter distance).


- Ride Duration
```{r}
ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=SNOW, y=tripduration, colour=usertype)) + geom_smooth() + labs(title = "Effects of Snow on Citi Bike Ride Duration", x = "Snow Depth (inches)", y = "Trip Duration (secs)")

ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=SNOW, y=tripduration, colour=usertype)) + geom_point(alpha = 0.5) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Ride Duration", x = "Snow Depth (inches)", y = "Trip Duration (secs)")

ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=roundedSNOW, y=tripduration, colour=usertype)) + geom_point(alpha = 0.5) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Ride Duration", x = "Snow Depth (rounded inches)", y = "Trip Duration (secs)")

totalRides <- tapply(masterdata$tripduration, masterdata$SNOW, mean, na.rm = TRUE)
barplot(totalRides)

totalRides <- tapply(masterdata$tripduration, masterdata$roundedSNOW, mean, na.rm = TRUE)
barplot(totalRides)
```

Trip duration drastically decreases when there is snow on the ground this is likely due to a combination of temperature and safety concerns. However, we see a much steeper drop off in ridership and trip duration among regular customers than subscribers.

A potential solution to increase business would be to increase incentives for non-subscribers to ride when their is snow (is pricing a concern for them?) >> risk: liability?


- Total Rides
```{r}
# Count - Raw Data
ggplot(data=masterdata, aes(x=SNOW)) + geom_bar()
ggplot(data=masterdata[masterdata$SNOW > 0, ], aes(x=SNOW)) + geom_bar()

# Count - Rounded Values
ggplot(data=masterdata, aes(x=roundedSNOW)) + geom_bar()
ggplot(data=masterdata[masterdata$roundedSNOW > 0, ], aes(x=roundedSNOW)) + geom_bar()

# Average - Rounded Values
masterdata %>%
  group_by(roundedSNOW) %>%
  summarise(
    numRides = mean(n())
  ) %>%
  ggplot(aes(x = roundedSNOW, y = numRides)) + geom_point() + ylim(0,1000) + labs(x = "Snow Depth", y = "Average Number of CitiBike Rides")
```

As expected the more snow there is on the ground the less riders (on average per day with that amount of snow) there are. The second plots doesn't include 0 values to remove days where there is no snow on the ground.


- Demographics (Female vs. Male, Age)
```{r}
ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=SNOW, y=tripduration, colour=gender)) + geom_smooth() + labs(title = "Effects of Snow on Citi Bike Riders based on Gender", x = "Snow Depth (inches)", y = "Trip Duration (secs)")

ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=SNOW, y=tripduration, colour=gender)) + geom_point(alpha = 0.25) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Riders based on Gender", x = "Snow Depth (inches)", y = "Trip Duration (secs)")

ggplot(data=masterdata[masterdata$tripduration < 10000,], aes(x=roundedSNOW, y=tripduration, colour=gender)) + geom_point(alpha = 0.25) + geom_jitter() + labs(title = "Effects of Snow on Citi Bike Riders based on Gender", x = "Snow Depth (rounded inches)", y = "Trip Duration (secs)")
```

It seems that males are more likely to continue riding CitiBikes when there is snow on the ground than Females or Unknowns (which is expected as Unknowns fall largely in the non-subscriber category), but the difference is minimal.


- Speed
```{r}
ggplot(data=masterdata, aes(x=SNOW, y=speedMetersperMin, colour=usertype)) + geom_smooth() + labs(title = "Effects of Snow on Citi Bike Rider Speed", x = "Snow Depth (inches)", y = "Speed (meters per sec)")
```

Subscribers are trying to get from point A to point B and actually go faster with more snow (likely to get out of the cold and we can assume NYC streets are well plowed despite indicated snow depth)

